---
title: "Homework - Regularizations"
author: "Jean Alvarez"
date: today
format:
  html:
    embed-resources: true
    toc: true
    number-sections: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

# Part 1: Iowa House Price Data — Lab Walkthrough Replication

## Libraries

```{r}
library(tidyverse)
library(h2o)
library(rsample)
library(yardstick)
library(skimr)
library(plotly)
library(readxl)
```

## Initiate H2O

```{r}
#| output: false
h2o.init()
h2o.no_progress()
```

## Load & Prep Iowa Data

Load the Iowa house price data. `SalePrice` is the target variable.

```{r}
iowa <- read_excel("ml_data/Original_Data.xlsx", skip = 3, na = "NA")

dim(iowa)
```

Drop fully-empty columns, drop `Id`, relocate the predicted variable to the left:

```{r}
iowa <- iowa |>
  select(where(\(x) !all(is.na(x)))) |>
  select(-Id) |>
  relocate(SalePrice)

iowa |> glimpse()
```

## Skim Data

```{r}
skim(iowa)
```

### Identify Columns with Too Many NAs

Drop variables with more than 20% missing values:

```{r}
too_many_NAs <- iowa |>
  select(where(\(x) sum(is.na(x)) / length(x) > 0.2)) |>
  colnames()

print(too_many_NAs)
```

## Convert Character to Factor

H2O treats factor variables as enumerations (`enum`) and handles one-hot encoding internally.

```{r}
iowa <- iowa |>
  mutate(across(where(is.character), as.factor))

glimpse(iowa)
```

## Predicted / Predictor Variables

```{r}
X <- iowa |>
  select(-all_of(too_many_NAs)) |>
  select(-SalePrice) |>
  colnames()

length(X)

Y <- "SalePrice"
```

## Sampling Data

Random train / validation / test split (70 / 15 / 15):

```{r}
set.seed(42)
split <- initial_validation_split(iowa, prop = c(0.7, 0.15))

train_h2o <- training(split)   |> as.h2o()
valid_h2o <- validation(split) |> as.h2o()
test_h2o  <- testing(split)    |> as.h2o()

h2o.describe(train_h2o)
```

## Hyperparameter Tuning

### Lasso (L1)

For Lasso, `lambda` is the only hyperparameter to tune. Grid search over
$\lambda \in [0, 500]$ in steps of 5.

```{r}
#| output: false
lambda_search <- list(lambda = seq(0, 500, by = 5))

lasso_grid <- h2o.grid(
  algorithm       = "glm",
  grid_id         = "lasso_grid",
  x               = X,
  y               = Y,
  training_frame  = train_h2o,
  validation_frame = valid_h2o,
  family          = "gaussian",
  alpha           = 1,          # Lasso
  standardize     = TRUE,
  seed            = 42,
  hyper_params    = lambda_search
)
```

Tuning result — best model by $R^2$:

```{r}
leaderboard_lasso_r2 <- h2o.getGrid(
  "lasso_grid",
  sort_by    = "r2",
  decreasing = TRUE
)

leaderboard_lasso_r2@summary_table |> as_tibble() |> head(10)
```

Extract the best Lasso model:

```{r}
best_lasso_model <- h2o.getModel(leaderboard_lasso_r2@model_ids[[1]])
best_lasso_model@parameters$lambda
```

Test-set performance:

```{r}
h2o.performance(best_lasso_model, newdata = test_h2o)
```

### Ridge (L2)

To tune a Ridge model, set `alpha = 0` and reuse the same lambda search space.

```{r}
#| output: false
ridge_grid <- h2o.grid(
  algorithm       = "glm",
  grid_id         = "ridge_grid",
  x               = X,
  y               = Y,
  training_frame  = train_h2o,
  validation_frame = valid_h2o,
  family          = "gaussian",
  alpha           = 0,          # Ridge
  standardize     = TRUE,
  seed            = 42,
  hyper_params    = lambda_search
)
```

```{r}
leaderboard_ridge_r2 <- h2o.getGrid(
  "ridge_grid",
  sort_by    = "r2",
  decreasing = TRUE
)

best_ridge_model <- h2o.getModel(leaderboard_ridge_r2@model_ids[[1]])
best_ridge_model@parameters$lambda
```

```{r}
h2o.performance(best_ridge_model, newdata = test_h2o)
```

### Elastic Net (L1 + L2)

Elastic Net mixes Lasso and Ridge. Since we now tune both $\alpha$ and $\lambda$,
the grid grows quickly — we switch to a **Random Discrete** search capped at 300 models.

```{r}
#| output: false
elastic_search <- list(
  alpha  = seq(0, 1, by = 0.1),      # L1 ratio
  lambda = seq(0, 500, by = 5)        # penalization
)

search_list <- list(
  strategy   = "RandomDiscrete",
  max_models = 300,
  seed       = 42
)

elastic_grid <- h2o.grid(
  algorithm        = "glm",
  grid_id          = "elastic_grid",
  x                = X,
  y                = Y,
  training_frame   = train_h2o,
  validation_frame = valid_h2o,
  family           = "gaussian",
  standardize      = TRUE,
  hyper_params     = elastic_search,
  search_criteria  = search_list
)
```

```{r}
leaderboard_elastic_r2 <- h2o.getGrid(
  "elastic_grid",
  sort_by    = "r2",
  decreasing = TRUE
)

best_elastic_model <- h2o.getModel(leaderboard_elastic_r2@model_ids[[1]])

list(
  alpha  = best_elastic_model@parameters$alpha,
  lambda = best_elastic_model@parameters$lambda
)
```

```{r}
h2o.performance(best_elastic_model, newdata = test_h2o)
```

### Baseline OLS (No Regularization)

```{r}
#| output: false
plain_ols_model <- h2o.glm(
  model_id         = "plain_ols_model",
  x                = X,
  y                = Y,
  training_frame   = train_h2o,
  validation_frame = valid_h2o,
  family           = "gaussian",
  lambda           = 0,    # no penalty
  compute_p_values = FALSE
)
```

```{r}
h2o.performance(plain_ols_model, newdata = test_h2o)
```

## Model Predictions

Attach predicted columns from every model onto the validation and test frames.

```{r}
#| output: false
make_predictions <- function(model_name) {
  colname <- str_glue("predicted_{model_name}")
  model   <- get(model_name, envir = .GlobalEnv)

  valid_h2o[colname] <<- h2o.predict(model, valid_h2o)
  test_h2o[colname]  <<- h2o.predict(model, test_h2o)
}

walk(
  c("best_lasso_model", "best_ridge_model",
    "best_elastic_model", "plain_ols_model"),
  make_predictions
)
```

## Tuning Visualizations

### Ridge — Lambda vs R² (2D)

```{r}
ridge_summary <- leaderboard_ridge_r2@summary_table |>
  as_tibble() |>
  mutate(lambda = as.numeric(lambda),
         r2     = as.numeric(r2))

ridge_summary |>
  ggplot(aes(x = lambda, y = r2)) +
  geom_point(color = "steelblue") +
  geom_line(color = "steelblue") +
  theme_bw() +
  scale_x_continuous(n.breaks = 10) +
  labs(title = "Ridge: R² vs Lambda",
       x = "Lambda (penalty)", y = "R²")
```

### Elastic Net — Lambda × Alpha × R² (3D)

```{r}
elastic_summary <- leaderboard_elastic_r2@summary_table |>
  as_tibble() |>
  mutate(alpha  = as.numeric(alpha),
         lambda = as.numeric(lambda),
         r2     = as.numeric(r2))

plot_ly(
  data = elastic_summary,
  x = ~lambda, y = ~alpha, z = ~r2,
  type = "scatter3d", mode = "markers",
  marker = list(size = 5, color = ~r2,
                colorscale = "Viridis", opacity = 0.8)
) |>
  layout(scene = list(
    xaxis = list(title = "Lambda (Penalty)"),
    yaxis = list(title = "Alpha (Mixture)"),
    zaxis = list(title = "R²")
  ))
```

## Prediction Visualizations

Copy predictions and truth back to R for plotting and metric evaluation.

```{r}
cols <- c("SalePrice",
          "predicted_best_lasso_model",
          "predicted_best_ridge_model",
          "predicted_best_elastic_model",
          "predicted_plain_ols_model")

valid_r <- as.data.frame(valid_h2o[, cols]) |> as_tibble()
test_r  <- as.data.frame(test_h2o[,  cols]) |> as_tibble()
```

### Actual vs Predicted (Validation Set)

```{r}
valid_r |>
  pivot_longer(cols = !SalePrice) |>
  mutate(name = fct_recode(
    name,
    Lasso   = "predicted_best_lasso_model",
    Ridge   = "predicted_best_ridge_model",
    Elastic = "predicted_best_elastic_model",
    OLS     = "predicted_plain_ols_model"
  )) |>
  ggplot(aes(x = value, y = SalePrice, color = name)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0,
              color = "red", linetype = "dotted") +
  facet_grid(cols = vars(name)) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(title = "Actual vs Predicted — Validation Set",
       x = "Predicted", y = "Actual SalePrice")
```

## Metric Comparisons

Evaluate each model on the held-out **test** set.

```{r}
my_metric <- metric_set(mae, rmse, mape, rsq_trad)

lasso_metrics <- my_metric(test_r, truth = SalePrice,
                           estimate = predicted_best_lasso_model) |>
  mutate(model = "Lasso")

ridge_metrics <- my_metric(test_r, truth = SalePrice,
                           estimate = predicted_best_ridge_model) |>
  mutate(model = "Ridge")

elastic_metrics <- my_metric(test_r, truth = SalePrice,
                             estimate = predicted_best_elastic_model) |>
  mutate(model = "Elastic Net")

ols_metrics <- my_metric(test_r, truth = SalePrice,
                         estimate = predicted_plain_ols_model) |>
  mutate(model = "OLS (Baseline)")

iowa_comparison <- bind_rows(
  ols_metrics, lasso_metrics, ridge_metrics, elastic_metrics
) |>
  select(model, .metric, .estimate) |>
  pivot_wider(names_from = .metric, values_from = .estimate)

iowa_comparison
```

### Performance Gain over OLS

```{r}
ols_rmse <- iowa_comparison |> filter(model == "OLS (Baseline)") |> pull(rmse)
ols_mae  <- iowa_comparison |> filter(model == "OLS (Baseline)") |> pull(mae)

iowa_comparison |>
  mutate(
    rmse_improvement_pct = (ols_rmse - rmse) / ols_rmse * 100,
    mae_improvement_pct  = (ols_mae  - mae)  / ols_mae  * 100
  ) |>
  select(model, rmse, rmse_improvement_pct,
         mae, mae_improvement_pct, rsq_trad)
```

## Variable Importance

In linear models, variable importance is the absolute value of each standardized coefficient.

```{r}
h2o.varimp_plot(best_lasso_model, num_of_features = 20)
```

```{r}
# How many variables did Lasso keep vs how many did it shrink to zero?
varimp_lasso <- h2o.varimp(best_lasso_model) |> as_tibble()

varimp_lasso |>
  summarise(
    total_features = n(),
    features_kept  = sum(coefficients != 0),
    features_zeroed = sum(coefficients == 0)
  )
```


# Part 2: MSFT AR(3)X Model with Regularization

From the previous lab, I built an AR(3)X model on MSFT monthly returns using
Fama-French 3 factors (lag 1–3 of `ret_m`, and lag 2–3 of `mkt_rf`, `smb`, `hml`).
Here I re-run that baseline OLS and then apply Lasso, Ridge, and Elastic Net
to see whether regularization improves out-of-sample performance on financial data.

## Libraries

```{r}
library(tidyquant)
library(curl)
```

## Download & Prep MSFT + FF3 Data

```{r}
msft_prices <- tq_get("MSFT", from = "2012-01-01", to = "2023-01-01")

msft_monthly <- msft_prices |>
  mutate(ret_d = adjusted / lag(adjusted) - 1) |>
  mutate(date  = floor_date(date, "month")) |>
  group_by(symbol, date) |>
  summarise(ret_m = prod(1 + ret_d, na.rm = TRUE) - 1, .groups = "drop")
```

```{r}
ff_url      <- "https://mba.tuck.dartmouth.edu/pages/faculty/ken.french/ftp/F-F_Research_Data_Factors_CSV.zip"
temp_addr   <- tempfile(fileext = ".zip")
multi_download(ff_url, temp_addr)

ff_factors <- read_csv(temp_addr, skip = 3) |>
  janitor::clean_names() |>
  drop_na() |>
  mutate(
    date   = ymd(paste0(x1, "01")),
    mkt_rf = as.numeric(mkt_rf) / 100,
    smb    = as.numeric(smb)    / 100,
    hml    = as.numeric(hml)    / 100,
    rf     = as.numeric(rf)     / 100,
    x1     = NULL
  )

msft_monthly <- msft_monthly |>
  left_join(ff_factors, by = join_by(date)) |>
  arrange(date)
```

## Generate Lagged Features (AR(3)X)

```{r}
msft_monthly <- msft_monthly |>
  mutate(
    ret_m_lag1 = lag(ret_m, 1),
    ret_m_lag2 = lag(ret_m, 2),
    ret_m_lag3 = lag(ret_m, 3),
    mkt_rf_lag2 = lag(mkt_rf, 2),
    mkt_rf_lag3 = lag(mkt_rf, 3),
    smb_lag2    = lag(smb, 2),
    smb_lag3    = lag(smb, 3),
    hml_lag2    = lag(hml, 2),
    hml_lag3    = lag(hml, 3)
  ) |>
  drop_na()

X_msft <- msft_monthly |> select(contains("lag")) |> colnames()
X_msft
```

## Temporal Train / Valid / Test Split

Since this is time-series data, we split by **time**, not randomly:
last 1 year → test, 2 years before that → validation, everything else → train.

```{r}
msft_monthly <- msft_monthly |> arrange(date)

cutoff_valid <- max(msft_monthly$date) - years(2)
cutoff_test  <- max(msft_monthly$date) - years(1)

train_msft <- msft_monthly |> filter(date <  cutoff_valid) |> as.h2o()
valid_msft <- msft_monthly |> filter(date >= cutoff_valid & date < cutoff_test) |> as.h2o()
test_msft  <- msft_monthly |> filter(date >= cutoff_test)  |> as.h2o()
```

## Baseline OLS (from previous lab)

```{r}
#| output: false
msft_ols <- h2o.glm(
  model_id         = "msft_ols",
  x                = X_msft,
  y                = "ret_m",
  training_frame   = train_msft,
  validation_frame = valid_msft,
  family           = "gaussian",
  lambda           = 0,
  standardize      = TRUE
)
```

```{r}
h2o.performance(msft_ols, newdata = test_msft)
```

## Lasso Tuning

Smaller lambda grid — monthly returns are on a tiny scale (~1% magnitude) so large
penalties wipe out all signal.

```{r}
#| output: false
msft_lambda_search <- list(lambda = seq(0, 0.05, by = 0.001))

msft_lasso_grid <- h2o.grid(
  algorithm        = "glm",
  grid_id          = "msft_lasso_grid",
  x                = X_msft,
  y                = "ret_m",
  training_frame   = train_msft,
  validation_frame = valid_msft,
  family           = "gaussian",
  alpha            = 1,
  standardize      = TRUE,
  seed             = 42,
  hyper_params     = msft_lambda_search
)
```

```{r}
msft_lasso_board <- h2o.getGrid("msft_lasso_grid",
                                sort_by = "r2", decreasing = TRUE)

best_msft_lasso <- h2o.getModel(msft_lasso_board@model_ids[[1]])
best_msft_lasso@parameters$lambda
```

## Ridge Tuning

```{r}
#| output: false
msft_ridge_grid <- h2o.grid(
  algorithm        = "glm",
  grid_id          = "msft_ridge_grid",
  x                = X_msft,
  y                = "ret_m",
  training_frame   = train_msft,
  validation_frame = valid_msft,
  family           = "gaussian",
  alpha            = 0,
  standardize      = TRUE,
  seed             = 42,
  hyper_params     = msft_lambda_search
)
```

```{r}
msft_ridge_board <- h2o.getGrid("msft_ridge_grid",
                                sort_by = "r2", decreasing = TRUE)

best_msft_ridge <- h2o.getModel(msft_ridge_board@model_ids[[1]])
best_msft_ridge@parameters$lambda
```

## Elastic Net Tuning

```{r}
#| output: false
msft_elastic_search <- list(
  alpha  = seq(0, 1, by = 0.1),
  lambda = seq(0, 0.05, by = 0.001)
)

msft_elastic_grid <- h2o.grid(
  algorithm        = "glm",
  grid_id          = "msft_elastic_grid",
  x                = X_msft,
  y                = "ret_m",
  training_frame   = train_msft,
  validation_frame = valid_msft,
  family           = "gaussian",
  standardize      = TRUE,
  hyper_params     = msft_elastic_search,
  search_criteria  = list(strategy = "RandomDiscrete",
                          max_models = 300,
                          seed = 42)
)
```

```{r}
msft_elastic_board <- h2o.getGrid("msft_elastic_grid",
                                  sort_by = "r2", decreasing = TRUE)

best_msft_elastic <- h2o.getModel(msft_elastic_board@model_ids[[1]])

list(
  alpha  = best_msft_elastic@parameters$alpha,
  lambda = best_msft_elastic@parameters$lambda
)
```

## Attach Predictions to Test Set

```{r}
#| output: false
test_msft[["predicted_ols"]]     <- h2o.predict(msft_ols,         test_msft)
test_msft[["predicted_lasso"]]   <- h2o.predict(best_msft_lasso,   test_msft)
test_msft[["predicted_ridge"]]   <- h2o.predict(best_msft_ridge,   test_msft)
test_msft[["predicted_elastic"]] <- h2o.predict(best_msft_elastic, test_msft)
```

```{r}
msft_test_r <- as.data.frame(test_msft[, c("date", "ret_m",
                                           "predicted_ols",
                                           "predicted_lasso",
                                           "predicted_ridge",
                                           "predicted_elastic")]) |>
  as_tibble()

msft_test_r |> head()
```

## Tuning Visualization — Ridge (λ vs R²)

```{r}
msft_ridge_summary <- msft_ridge_board@summary_table |>
  as_tibble() |>
  mutate(lambda = as.numeric(lambda),
         r2     = as.numeric(r2))

msft_ridge_summary |>
  ggplot(aes(x = lambda, y = r2)) +
  geom_point(color = "darkgreen") +
  geom_line(color = "darkgreen") +
  theme_bw() +
  labs(title = "MSFT Ridge: R² vs Lambda",
       x = "Lambda (penalty)", y = "R² (validation)")
```

## Tuning Visualization — Elastic Net (α × λ × R²)

```{r}
msft_elastic_summary <- msft_elastic_board@summary_table |>
  as_tibble() |>
  mutate(alpha  = as.numeric(alpha),
         lambda = as.numeric(lambda),
         r2     = as.numeric(r2))

plot_ly(
  data = msft_elastic_summary,
  x = ~lambda, y = ~alpha, z = ~r2,
  type = "scatter3d", mode = "markers",
  marker = list(size = 5, color = ~r2,
                colorscale = "Viridis", opacity = 0.8)
) |>
  layout(scene = list(
    xaxis = list(title = "Lambda (Penalty)"),
    yaxis = list(title = "Alpha (Mixture)"),
    zaxis = list(title = "R²")
  ),
  title = "MSFT Elastic Net Tuning Surface")
```

## Actual vs Predicted — MSFT Test Set

```{r}
msft_test_r |>
  pivot_longer(cols = starts_with("predicted_"),
               names_to = "model", values_to = "predicted") |>
  mutate(model = fct_recode(model,
                            OLS        = "predicted_ols",
                            Lasso      = "predicted_lasso",
                            Ridge      = "predicted_ridge",
                            `Elastic`  = "predicted_elastic")) |>
  ggplot(aes(x = predicted, y = ret_m, color = model)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_abline(slope = 1, intercept = 0,
              color = "red", linetype = "dotted") +
  facet_grid(cols = vars(model)) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(title = "MSFT AR(3)X: Actual vs Predicted (Test Set)",
       x = "Predicted Monthly Return",
       y = "Actual Monthly Return")
```

## Metric Comparison — MSFT Models

```{r}
msft_metric <- metric_set(mae, rmse, rsq_trad)

msft_comparison <- bind_rows(
  msft_metric(msft_test_r, truth = ret_m, estimate = predicted_ols)     |> mutate(model = "OLS (Baseline)"),
  msft_metric(msft_test_r, truth = ret_m, estimate = predicted_lasso)   |> mutate(model = "Lasso"),
  msft_metric(msft_test_r, truth = ret_m, estimate = predicted_ridge)   |> mutate(model = "Ridge"),
  msft_metric(msft_test_r, truth = ret_m, estimate = predicted_elastic) |> mutate(model = "Elastic Net")
) |>
  select(model, .metric, .estimate) |>
  pivot_wider(names_from = .metric, values_from = .estimate)

msft_comparison
```

### Performance Gain Over OLS — MSFT

```{r}
msft_ols_rmse <- msft_comparison |> filter(model == "OLS (Baseline)") |> pull(rmse)
msft_ols_mae  <- msft_comparison |> filter(model == "OLS (Baseline)") |> pull(mae)

msft_comparison |>
  mutate(
    rmse_improvement_pct = (msft_ols_rmse - rmse) / msft_ols_rmse * 100,
    mae_improvement_pct  = (msft_ols_mae  - mae)  / msft_ols_mae  * 100
  ) |>
  select(model, rmse, rmse_improvement_pct,
         mae, mae_improvement_pct, rsq_trad)
```

## Variable Importance — MSFT Lasso

```{r}
h2o.varimp_plot(best_msft_lasso)
```

```{r}
h2o.varimp(best_msft_lasso) |>
  as_tibble() |>
  select(variable, relative_importance, scaled_importance,
         percentage, coefficients, standardized_coefficients)
```

# Discussion

**Iowa housing data (Part 1).** With 74 features — many of them one-hot-encoded
categorical levels — this is exactly the high-dimensional setting where
regularization shines. OLS overfits noisy neighborhood and categorical dummies,
while Lasso/Ridge/Elastic Net shrink or zero-out those unreliable coefficients.
The comparison table shows meaningful RMSE and MAE reductions relative to plain
OLS, and the Lasso variable-importance plot confirms the model automatically
performs feature selection (several coefficients driven to exactly zero).

**MSFT AR(3)X (Part 2).** Only 9 predictors, and stock returns are famously noisy
with a low signal-to-noise ratio. Regularization matters here too, but the gains
are smaller in absolute terms: Lasso typically zeros out several lag terms,
Ridge keeps all lags but shrinks them, and Elastic Net lands in between. Because
the best Lasso lambda tends to be very small (monthly returns are ~1% magnitude),
regularization mostly acts as a safety net against overfitting rather than a
dramatic accuracy booster. The takeaway that matters for financial forecasting:
regularization reduces variance of the coefficient estimates — making the model's
behavior more stable out-of-sample — even when test R² stays low.

```{r}
#| output: false
h2o.shutdown(prompt = FALSE)
```
